Efficient implementation of penalized regression\\for genetic risk prediction

### 2018-09-24 paper/PRS.tex
some reformating

@@

\section{Methods}

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.

@@

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.

@@

In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes. Note that a similar penalized linear regression is also available in our software. 
Usually, joint statistical learning methods are so computationally demanding that they cannot be used on large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}. 
A partial solution to this problem is to pre-filter data, but it usually leads to a decrease in predictive performance as some information is removed \cite[]{accuheight2018}. This pre-filtering step is not necessary anymore thanks to our implementation.
Moreover, the model we present in this paper can flexibly adapt to traits with different architectures thanks to two hyper-parameters that are automatically learned with the model. Thus, we strongly believe that our implementation of penalized regressions would be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height). 
Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.

@@

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & PRS \\
& & & 30 in all & & & & PLR \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & PRS \\ 
& & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome. Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

\clearpage
\section*{Acknowledgements}

@@

\section*{Reproducibility}

All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
Results are available at [TODO]. A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.

\vspace*{-12pt}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Supplementary Materials}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\subsection*{Model ``COMP''}

For model ``COMP'', we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.

\subsection*{Maximum AUCs} \label{sec:auc-max}

We used three different ways to estimate the maximum achievable AUC for our simulations.
First, we used the estimation from equation (3) of \cite{wray2010genetic}. For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%). Note that this approximation is reported to be less accurate for high heritabilities.
Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations. We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates. For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
Given all these estimates of maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the simulation parameters.

\begin{table}[ht]
\centering
\begin{tabular}{|l|cccc|r|}
\hline
Population & UK & Finland & Netherlands & Italy & Total\\
\hline
Cases & 2569 & 637 & 795 & 495 & 4496\\
Controls & 7492 & 1799 & 828 & 540 & 10659\\
\hline
Total & 10061 & 2436 & 1623 & 1035 & 15155\\
\hline
\end{tabular}
\caption{Number of individuals by population and disease status in the celiac disease case-control study (after quality control, genotyped on 281,122 SNPs).\label{tab:celiac-data}}
\end{table}

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Thu Oct 26 16:46:26 2017
\begin{table}[ht]
\centering
\begin{tabular}{ccccccccccccccc}
  \hline
1.00e+00 & 7.22e-01 & 5.87e-01 & 4.20e-01 & 2.43e-01 & 1.00e-01 & 2.35e-02 & 2.21e-03 & 4.69e-05 & 8.81e-08 & 3.18e-12 & 1.83e-19 & 2.89e-31 & 1.70e-50 & 7.71e-82 \\ 
  5.00e-08 & 7.05e-01 & 5.65e-01 & 3.95e-01 & 2.20e-01 & 8.47e-02 & 1.79e-02 & 1.42e-03 & 2.28e-05 & 2.73e-08 & 4.69e-13 & 8.08e-21 & 1.80e-33 & 4.30e-54 & 1.06e-87 \\ 
  7.94e-01 & 6.87e-01 & 5.42e-01 & 3.69e-01 & 1.97e-01 & 7.08e-02 & 1.34e-02 & 8.83e-04 & 1.05e-05 & 7.74e-09 & 6.03e-14 & 2.86e-22 & 7.73e-36 & 5.97e-58 & 5.49e-94 \\ 
  7.81e-01 & 6.69e-01 & 5.19e-01 & 3.43e-01 & 1.75e-01 & 5.85e-02 & 9.79e-03 & 5.31e-04 & 4.61e-06 & 2.01e-09 & 6.69e-15 & 7.92e-24 & 2.24e-38 & 4.37e-62 & 1.00e-100 \\ 
  7.67e-01 & 6.50e-01 & 4.95e-01 & 3.18e-01 & 1.54e-01 & 4.76e-02 & 7.01e-03 & 3.08e-04 & 1.90e-06 & 4.72e-10 & 6.32e-16 & 1.70e-25 & 4.26e-41 & 1.61e-66 &  \\ 
  7.53e-01 & 6.30e-01 & 4.70e-01 & 2.93e-01 & 1.35e-01 & 3.82e-02 & 4.90e-03 & 1.72e-04 & 7.31e-07 & 1.00e-10 & 5.04e-17 & 2.75e-27 & 5.16e-44 & 2.83e-71 &  \\ 
  7.38e-01 & 6.09e-01 & 4.46e-01 & 2.68e-01 & 1.17e-01 & 3.02e-02 & 3.33e-03 & 9.18e-05 & 2.63e-07 & 1.89e-11 & 3.35e-18 & 3.31e-29 & 3.84e-47 & 2.26e-76 &  \\ 
   \hline
\end{tabular}
\caption{The 102 thresholds used for the C+T method for this study.\label{tab:thr}}
\end{table}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{simple-CMSA}}
\caption{Illustration of one turn of the Cross-Model Selection and Averaging (CMSA) procedure. First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set (red) and the other ($K - 1$) folds form an inner training set (blue). A ``regularization path'' of models is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed. The model that minimizes the loss on the inner validation set is selected. Finally, the $K$ resulting models are averaged. 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.}
\label{fig:CMSA}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
\caption{Correlation between AUC and partial AUC values in scenario \textnumero1. There is a Spearman correlation of 98\% between values of AUC and partial AUC. The relation between the two values are the same whatever are the disease heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean of AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-logit}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean of AUC over 100 simulations. Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-PRS}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
\caption{Comparison of models in scenario \textnumero1 for model ``ADD''. Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-all-r2}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''. Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}). Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-chr6-all-r2}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
\caption{Comparison of T-Trees and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 5 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 5 simulations. \textbf{C:} Boxplots of execution times for 5 simulations.}
\label{fig:supp-ttrees}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-triple}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 50\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-triple}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
\caption{Comparison of PLR and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression -- \cite{zeng2017efficient}) in scenario \textnumero1. Simulations use model ``ADD'', an heritability of 80\% and $\alpha = 1$. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-biglasso}
\end{figure}

>>>>

\usepackage{xr}
\externaldocument{PRS-supp}

@@

\section{Material and Methods}

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.

@@

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.

@@

In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes. Note that a similar penalized linear regression is also available in our software. Our approach solves the dramatic computational burden faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.

We also demonstrated in simulations that our implementation of penalized regressions remains highly effective over a broad range of disease architectures. It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).  Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.

@@

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
& & & 30 in all & & & & PLR \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
& & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome. Mean of AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

\clearpage

\section*{Description of Supplemental Data}

Supplemental Data include a PDF with two sections of methods, two tables and ten figures. 
Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes.

\section*{Declaration of Interests}

The authors declare no competing interests.

\section*{Acknowledgements}

@@

\section*{Web Resources}

Results of simulations are available at [TODO]. A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}. The two R packages are available on GitHub.

@@
